---
title: "Notebook 8 — PD Policy Simulation (Logistic Regression + Realized Economics)"
author: "Ryan Owens"
date: "2/2/2026"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
execute:
  echo: true
  warning: false
  message: false
---

## Objective

Train an interpretable origination PD model (plain logistic regression) on mature vintages,
then simulate underwriting policies on a held-out year using **realized cashflow economics**.

### Economics (never used as model features)

If available (or computable from cashflow columns):

- **Realized net loss (principal net of recoveries)**  
  `realized_net_loss = loan_amnt - total_rec_prncp - recoveries + collection_recovery_fee`

- **Realized net profit (simple)**  
  `realized_net_profit = total_rec_int + total_rec_late_fee - realized_net_loss`

### Proxies (fallback only)

- **Loss proxy** (conservative): `loan_amnt * default_flag * LGD`
- **Revenue proxy** (simple): `installment * horizon_months`

For each policy: **"Decline top X% highest-PD loans"**, compute:

- Defaults avoided
- Loss avoided
- Profit change (vs baseline)
- Net impact (policy − baseline)

---

```{r libraries}
library(arrow)
library(dplyr)
library(tidyr)
library(janitor)
library(lubridate)
library(ggplot2)
library(pROC)
library(here)
library(scales)

```

```{r code}
# ----------------------------
# Paths
# ----------------------------
IN_PATH      <- here("data", "processed", "loans_model_table.parquet")
OUT_SCORED   <- here("data", "processed", "pd_scored_2016.parquet")
OUT_DECILES  <- here("data", "processed", "pd_deciles_2016.parquet")
OUT_POLICY   <- here("data", "processed", "pd_policy_curve_2016.parquet")

# ----------------------------
# Time split (mature backtest)
# ----------------------------
train_end  <- "2015-12"
test_start <- "2016-01"
test_end   <- "2016-12"

# ----------------------------
# Proxy knobs (fallback only)
# ----------------------------
LGD <- 1.0
horizon_months <- 12

# ----------------------------
# Model features (origination-only; no leakage)
# ----------------------------
feature_cols <- c(
  "fico_mid","dti","int_rate","loan_amnt","installment",
  "revol_util","open_acc","total_acc","inq_last_6mths","delinq_2yrs","pub_rec",
  "term","home_ownership","verification_status","purpose","addr_state"
)

# ----------------------------
# Economics columns (ex-post; never used in model)
# ----------------------------
econ_base_cols <- c(
  "loan_amnt",
  "total_rec_prncp","total_rec_int","total_rec_late_fee",
  "recoveries","collection_recovery_fee"
)

econ_optional_cols <- c(
  "total_pymnt","total_pymnt_inv",
  "realized_net_loss","realized_net_profit" # if you precomputed upstream
)
```


```{r code}

dt <- read_parquet(IN_PATH) |>
  clean_names() |>
  mutate(issue_ym = as.character(issue_ym))

cat("Rows:", nrow(dt), "\n")
cat("Terminal bad-outcome rate (completed-loans population):",
    round(mean(dt$default_flag, na.rm = TRUE), 4), "\n")
cat("Issue range:", min(dt$issue_ym, na.rm = TRUE), "→", max(dt$issue_ym, na.rm = TRUE), "\n")

missing_features <- setdiff(feature_cols, names(dt))
if (length(missing_features) > 0) {
  stop("Missing feature columns in loans_model_table.parquet: ",
       paste(missing_features, collapse = ", "))
}

present_optional <- intersect(econ_optional_cols, names(dt))
cat("Optional econ cols present:", paste(present_optional, collapse = ", "), "\n")


```

Train test split on before 2016 as 2017/18 data is too close to origination and not accurate

```{r tt split}
# Train: <= 2015-12
# Test : 2016-01 .. 2016-12
dt_train <- dt |> filter(issue_ym <= train_end)
dt_test  <- dt |> filter(issue_ym >= test_start, issue_ym <= test_end)

cat("Train rows:", nrow(dt_train),
    " Default rate:", round(mean(dt_train$default_flag, na.rm = TRUE), 4), "\n")
cat("Test rows :", nrow(dt_test),
    " Default rate:", round(mean(dt_test$default_flag,  na.rm = TRUE), 4), "\n")


```

```{r code}

# IMPORTANT:
# issue_ym is ONLY for splitting/inspection.
# Do NOT feed issue_ym into glm or predict to avoid factor-level/time leakage issues.

needed_model <- c("default_flag", feature_cols)

train_cc <- dt_train |>
  select(all_of(needed_model)) |>
  drop_na()

# For test, keep econ cols if present (or can be computed)
needed_test <- unique(c("issue_ym", needed_model, econ_base_cols, econ_optional_cols))
test_cc <- dt_test |>
  select(any_of(needed_test)) |>
  drop_na()

cat("Train complete-case:", nrow(train_cc), "\n")
cat("Test complete-case :", nrow(test_cc),  "\n")


```


```{r log reg}

model_glm <- glm(
  default_flag ~ .,
  data = train_cc,
  family = binomial(link = "logit")
)

summary(model_glm)

```



```{r roc_auc}

test_cc <- test_cc |>
  mutate(pd = predict(model_glm, newdata = test_cc, type = "response"))

roc_obj <- roc(test_cc$default_flag, test_cc$pd)
auc_val <- auc(roc_obj)

cat("Test AUC (2016):", round(as.numeric(auc_val), 4), "\n")

plot(
  roc_obj,
  lwd = 2,
  main = paste0("ROC — Logistic PD (Train≤", train_end, ", Test=2016)")
)
abline(a = 0, b = 1, lty = 2, col = "grey70")

```


```{r deciles}

# ------------------------------------------------------------------------------
# ECONOMICS: prefer realized, otherwise fall back to proxy
# ------------------------------------------------------------------------------
# We will end with:
#   realized_loss   (>=0)
#   realized_profit (can be negative)
# If those cannot be computed, we'll still compute proxies.

has_realized <- all(c("realized_net_loss", "realized_net_profit") %in% names(test_cc))

# If not precomputed, attempt to compute from base cashflow fields
if (!has_realized) {
  if (all(econ_base_cols %in% names(test_cc))) {
    test_cc <- test_cc |>
      mutate(
        realized_net_loss = as.numeric(loan_amnt) -
          as.numeric(total_rec_prncp) -
          as.numeric(recoveries) +
          as.numeric(collection_recovery_fee),

        # simple realized profit: interest + late fees minus realized net loss
        realized_net_profit =
          as.numeric(total_rec_int) +
          as.numeric(total_rec_late_fee) -
          realized_net_loss
      )
    has_realized <- TRUE
  }
}

# Standardize final realized fields we will actually use downstream:
if (has_realized) {
  test_cc <- test_cc |>
    mutate(
      realized_loss   = pmax(as.numeric(realized_net_loss), 0),
      realized_profit = as.numeric(realized_net_profit)
    )
} else {
  test_cc <- test_cc |>
    mutate(
      realized_loss   = NA_real_,
      realized_profit = NA_real_
    )
}

cat("Using realized economics:", has_realized, "\n")
```


```{r deciles}

scored_2016 <- test_cc |>
  mutate(
    # Proxies always computed (fallback + comparison)
    loss_proxy = loan_amnt * default_flag * LGD,
    rev_proxy  = installment * horizon_months,
    net_proxy  = rev_proxy - loss_proxy,

    # PD buckets
    risk_decile = ntile(pd, 10) # 10 = highest PD (riskiest)
  )

deciles_tbl <- scored_2016 |>
  group_by(risk_decile) |>
  summarise(
    loans        = n(),
    defaults     = sum(default_flag == 1, na.rm = TRUE),
    default_rate = mean(default_flag, na.rm = TRUE),
    avg_pd       = mean(pd, na.rm = TRUE),
    total_balance = sum(loan_amnt, na.rm = TRUE),

    # Realized economics (if available)
    realized_loss   = if (has_realized) sum(realized_loss, na.rm = TRUE) else NA_real_,
    realized_profit = if (has_realized) sum(realized_profit, na.rm = TRUE) else NA_real_,

    # Proxy economics
    loss_proxy = sum(loss_proxy, na.rm = TRUE),
    rev_proxy  = sum(rev_proxy,  na.rm = TRUE),
    net_proxy  = sum(net_proxy,  na.rm = TRUE),

    .groups = "drop"
  ) |>
  arrange(desc(risk_decile))

deciles_tbl


```


```{r output}
# ------------------------------------------------------------------------------
# WRITE OUTPUTS
# ------------------------------------------------------------------------------
write_parquet(scored_2016, OUT_SCORED)
write_parquet(deciles_tbl, OUT_DECILES)

cat("Saved:", OUT_SCORED, "\n")
cat("Saved:", OUT_DECILES, "\n")
```

```{r baseline}
baseline <- scored_2016 |>
  summarise(
    loans    = n(),
    defaults = sum(default_flag == 1, na.rm = TRUE),

    # realized
    realized_loss   = if (has_realized) sum(realized_loss, na.rm = TRUE) else NA_real_,
    realized_profit = if (has_realized) sum(realized_profit, na.rm = TRUE) else NA_real_,

    # proxy
    loss_proxy = sum(loss_proxy, na.rm = TRUE),
    rev_proxy  = sum(rev_proxy,  na.rm = TRUE),
    net_proxy  = sum(net_proxy,  na.rm = TRUE),

    .groups = "drop"
  )

baseline

```


```{r plot}
# ------------------------------------------------------------------------------
# PLOTS
# ------------------------------------------------------------------------------
# Chart 1: Default rate by PD decile (10 = highest PD)
p1 <- ggplot(deciles_tbl, aes(x = factor(risk_decile), y = default_rate)) +
  geom_col() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Held-out Year (2016) — Observed Default Rate by PD Decile",
    x = "Risk decile (10 = highest PD)",
    y = "Observed default rate"
  ) +
  theme_minimal(base_size = 12)

print(p1)

# Chart 2: Net impact vs decline %
p2 <- ggplot(policy_tbl, aes(x = decline_pct, y = net_impact)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Policy Simulation (Realized) — Net Impact vs Declining Top Risk",
    x = "Decline top X% by PD",
    y = "Net impact (policy − baseline)"
  ) +
  theme_minimal(base_size = 12)

print(p2)
```

```{r curve}
policy_curve <- function(df, decline_pct) {
  stopifnot(decline_pct >= 0, decline_pct <= 1)

  df <- df |> arrange(desc(pd))
  n_total   <- nrow(df)
  n_decline <- floor(n_total * decline_pct)

  declined <- if (n_decline > 0) df[1:n_decline, ] else df[0, ]
  kept     <- if (n_decline > 0) df[(n_decline + 1):n_total, ] else df

  use_realized <- has_realized && !all(is.na(df$realized_profit))

  if (use_realized) {
    kept_s <- kept |>
      summarise(
        kept_n        = n(),
        kept_defaults = sum(default_flag == 1, na.rm = TRUE),
        kept_loss     = sum(realized_loss, na.rm = TRUE),
        kept_profit   = sum(realized_profit, na.rm = TRUE),
        .groups = "drop"
      )

    tibble(
      decline_pct = decline_pct,
      decline_n   = n_decline,
      kept_n      = kept_s$kept_n,

      defaults_baseline = baseline$defaults,
      defaults_kept     = kept_s$kept_defaults,
      defaults_avoided  = baseline$defaults - kept_s$kept_defaults,

      loss_baseline  = baseline$realized_loss,
      loss_kept      = kept_s$kept_loss,
      loss_avoided   = baseline$realized_loss - kept_s$kept_loss,

      profit_baseline = baseline$realized_profit,
      profit_kept     = kept_s$kept_profit,
      profit_change   = kept_s$kept_profit - baseline$realized_profit,

      net_baseline = baseline$realized_profit,
      net_kept     = kept_s$kept_profit,
      net_impact   = kept_s$kept_profit - baseline$realized_profit,

      econ_mode = "realized"
    )
  } else {
    kept_s <- kept |>
      summarise(
        kept_n        = n(),
        kept_defaults = sum(default_flag == 1, na.rm = TRUE),
        kept_loss     = sum(loss_proxy, na.rm = TRUE),
        kept_rev      = sum(rev_proxy,  na.rm = TRUE),
        kept_net      = sum(net_proxy,  na.rm = TRUE),
        .groups = "drop"
      )

    tibble(
      decline_pct = decline_pct,
      decline_n   = n_decline,
      kept_n      = kept_s$kept_n,

      defaults_baseline = baseline$defaults,
      defaults_kept     = kept_s$kept_defaults,
      defaults_avoided  = baseline$defaults - kept_s$kept_defaults,

      loss_baseline = baseline$loss_proxy,
      loss_kept     = kept_s$kept_loss,
      loss_avoided  = baseline$loss_proxy - kept_s$kept_loss,

      rev_baseline = baseline$rev_proxy,
      rev_kept     = kept_s$kept_rev,
      rev_forgone  = baseline$rev_proxy - kept_s$kept_rev,

      net_baseline = baseline$net_proxy,
      net_kept     = kept_s$kept_net,
      net_impact   = kept_s$kept_net - baseline$net_proxy,

      econ_mode = "proxy"
    )
  }
}

grid <- seq(0, 0.30, by = 0.01)
policy_tbl <- bind_rows(lapply(grid, function(x) policy_curve(scored_2016, x)))

policy_tbl |> slice_head(n = 12)

write_parquet(policy_tbl, OUT_POLICY)
cat("Saved:", OUT_POLICY, "\n")

```

```{r chart}
# Chart 1: Default rate by PD decile
p1 <- ggplot(deciles_tbl, aes(x = factor(risk_decile), y = default_rate)) +
  geom_col() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Held-out Year (2016) — Observed Default Rate by PD Decile",
    x = "Risk decile (10 = highest PD)",
    y = "Observed default rate"
  ) +
  theme_minimal(base_size = 12)

print(p1)

# Chart 2: Net impact vs decline %
p2 <- ggplot(policy_tbl, aes(x = decline_pct, y = net_impact)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = paste0("Policy Simulation (", unique(policy_tbl$econ_mode), ") — Net Impact vs Declining Top Risk"),
    x = "Decline top X% by PD",
    y = "Net impact (policy − baseline)"
  ) +
  theme_minimal(base_size = 12)

print(p2)

```